---
title: "Creating a BDS Finding ADaM - ADLB"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("pharmaverse/admiral.test", ref = "main") # This is a required dependency of {admiral}
remotes::install_github("pharmaverse/admiral", ref = "main")
```

# Introduction

This article describes creating a BDS finding ADaM. Examples are currently 
presented and tested in the context of ADVS.  However, the examples could be
applied to other BDS Finding ADaMs such as ADEG, ADLB, etc. where a single 
result is captured in an SDTM Finding domain on a single date and/or time.

**Note**: *All examples assume CDISC SDTM and/or ADaM format as input unless 
otherwise specified.*

# Programming Workflow

* [Read in Data](#readdata)
* [Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`)](#datetime)
* [Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1`](#paramcd)
* [Derive Results (`AVAL`, `AVALC`)](#aval)
* [Derive Additional Parameters (e.g. `BSA`, `BMI`, or `MAP` for `ADVS`)](#derive_param)
* [Derive Timing Variables (e.g. `APHASE`, `AVISIT`, `APERIOD`)](#timing)
* [Timing Flag Variables (e.g. `ONTRTFL`)](#timingflag)
* [Assign Reference Range Indicator (`ANRIND`)](#referencerange)
* [Derive Baseline (`BASETYPE`, `ABLFL`, `BASE`, `BASEC`, `BNRIND`)](#baseline)
* [Derive Change from Baseline (`CHG`, `PCHG`)](#bchange)
* [Derive Analysis Flags (e.g. `ANL01FL`)](#analysisrec)
* [Assign Treatment (`TRTA`, `TRTP`)](#treatment)
* [Assign `ASEQ`](#aseq)
* [Derive Categorization Variables (`AVALCATx`)](#cat)
* [Add ADSL variables](#adsl_vars)
* [Derive New Rows](#additional)

## Read in Data {#readdata}

```{r}
library(admiral)
library(dplyr)
library(admiral.test)
library(lubridate)
library(stringr)
library(tibble)


data("adsl")
data("lb")
lb <- convert_blanks_to_na(lb) # Maybe explain why we do this
```

```{r}
adsl_vars <- vars(TRTSDT, TRTEDT, TRT01A, TRT01P)

adsl_lb <- left_join(
  lb,
  select(adsl, STUDYID, USUBJID, !!!adsl_vars),
  by = c("STUDYID", "USUBJID")
)
```


 
```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl_lb, 
  display_vars = vars(USUBJID, LBTESTCD, LBDTC, VISIT, TRTSDT, TRTEDT, TRT01A, TRT01P),
    filter = LBTESTCD == "ALB" & VISIT == "WEEK 2"
)
```

## Derive/Impute Numeric Date/Time and Analysis Day (`ADT`, `ADTM`, `ADY`, `ADTF`, `ATMF`) {#datetime}

The function `derive_vars_dt()` can be used to derive `ADT`. This function allows 
the user to impute the date as well.

Example calls:

```{r eval=TRUE}

# Should we introduce some partial dates here to show off impuration and preserve functions?

adsl_lb_dt <- derive_vars_dt(adsl_lb, new_vars_prefix = "A", dtc = LBDTC)
```


```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl_lb_dt, 
  display_vars = vars(USUBJID, VISIT, LBDTC, ADT, ADTF),
  filter = LBTESTCD == "ALB"
)
```

Similarly, `ADTM` may be created using the function `derive_vars_dtm()`. 
Imputation may be done on both the date and time components of `ADTM`.

```{r eval=FALSE}
# CDISC Pilot data does not contain times and the output of the derivation
# ADTM is not presented.
adsl_lb_dtm <- derive_vars_dtm(
  adsl_lb, 
  new_vars_prefix = "A", 
  dtc = LBDTC, 
  date_imputation = "FIRST"
)
```


```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl_lb_dtm, 
  display_vars = vars(USUBJID, VISIT, LBDTC, ADTM, ADTF),
  filter = LBTESTCD == "ALB"
)

```

```{r}
adsl_lb_ady <- derive_var_ady(adsl_lb_dtm, reference_date = TRTSDT, date = ADTM)

```


```{r, eval=TRUE, echo=FALSE}
dataset_vignette(
  adsl_lb_ady, 
  display_vars = vars(USUBJID, VISIT, TRTSDT, ADTM, ADY,),
  filter = USUBJID == "01-716-1024"
)
```


## Assign `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1` {#paramcd}

To assign parameter level values such as `PARAMCD`, `PARAM`, `PARAMN`, `PARCAT1`,
etc., a lookup can be created to join to the source data.

For example, when creating `ADVS`, a lookup based on the SDTM `--TESTCD` value 
may be created:

```{r, eval=TRUE, echo=FALSE}
param_lookup <- lb %>% count(LBTEST, LBTESTCD, LBORRESU, LBCAT) %>% 
  mutate(PARAM = str_c(LBTEST, " (", LBORRESU, ")"),
         PARAMCD = LBTESTCD,
         PARCAT1 = str_to_sentence(LBCAT),
         PARAMN = seq(1:length(PARAMCD))) %>% 
  select(-n)

```


```{r eval=TRUE}
adsl_lb_pcd <- left_join(adsl_lb_ady, select(param_lookup, LBTESTCD, PARAMCD), by = "LBTESTCD")
```


## Derive Timing Variables (e.g. `APHASE`, `AVISIT`, `APERIOD`) {#timing}

Categorical timing variables are protocol and analysis dependent.  Below is a 
simple example. 

